---
title: "Cost effectiveness analysis for an AMR point-of-care test"
author: "Adam Howes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

* Neonate with suspected neonatal sepsis in India
* Considered two possible point-of-care tests (POCT)
  a. POCT to identify neonatal sepsis, but not causative pathogen **modelled in Scenario 2**
  b. POCT to identify neonatal sepsis and causative pathogen **not yet modelled**
* Metrics are cost per disability adjusted life year (DALY) and incremental cost-effectiveness ratio (ICER)

```{r}
#' Number of children born in India each year
births_per_year <- 25 * 10^6

#' Rate of having neonatal sepsis
sepsis_rate <- 70 / 1000

#' The number of births with neonatal sepsis is
(births_sepsis <- births_per_year * sepsis_rate)

#' Of those, assume 40% are high risk and 60% are low risk
prop_high_risk_sepsis <- 0.4
prop_low_risk_sepsis <- 1 - prop_high_risk_sepsis

#' Total births with high risk sepsis
#' Assume that all high risk cases recive correct treatment
#' TODO Why? Are they easier to spot somehow?
(births_sepsis_high_risk <- births_sepsis * prop_high_risk_sepsis)

#' Total births with low risk sepsis
(births_sepsis_low_risk <- births_sepsis * prop_low_risk_sepsis)

#' Assume 20% of neonates receive antibiotics
prop_abx <- 0.2

#' Then the total number of births who receive anitbiotics per year is
(births_abx <- births_per_year * prop_abx)

#' Assume that all high risk births receive antibiotics
(births_abx_sepsis_high_risk <- births_sepsis_high_risk)

#' Assume the remainder are given to those who are low risk
(births_abx_sepsis_low_risk <- births_abx - births_abx_sepsis_high_risk)
```

```{r}
#' AMR burden per prescription of antibiotics
total_births_per_year <- 130 * 10^6
(total_births_abx_per_year <- total_births_per_year * prop_abx)
total_sepsis_burden <- 24.6 * 10^6
prop_burden_abx <- 0.31
total_amr_burden <- total_sepsis_burden * prop_burden_abx
(amr_burden_per_abx <- total_amr_burden / total_births_abx_per_year)
```

```{r}
cost_culture <- 50
total_cost_rdt <- 40
subsidy_rdt <- 20
(cost_rdt <- total_cost_rdt - subsidy_rdt)
cost_treatment <- 10
```

## Scenario 1

```{r}
#' Number of DALYs averted by correctly treating
benefit_correct_sepsis_diagnosis <- 9

#' Treat every neonate with low risk
benefit_empiric <- benefit_correct_sepsis_diagnosis * births_sepsis_low_risk - amr_burden_per_abx * (births_abx_sepsis_low_risk - births_sepsis_low_risk)

#' Total cost to culture and treat all
cost_empiric <- (cost_culture + cost_treatment) * births_abx_sepsis_low_risk

#' Cost per DALY
(cost_per_daly_s1 <- cost_empiric / benefit_empiric)
```

## Scenario 2

$$
\text{sensitivity} = \frac{\text{TP}}{{\text{TP} + \text{FN}}}
$$
$$
\text{specificity} = \frac{\text{TN}}{{\text{TN} + \text{FP}}}
$$

```{r}
sensitivity <- 0.9
specificity <- 0.9

sepsis_prev_low_risk <- births_sepsis_low_risk / births_abx_sepsis_low_risk

#' Has sepsis and treated
(true_positive <- births_abx_sepsis_low_risk * sepsis_prev_low_risk * sensitivity)
  
#' Doesn't have sepsis and treated
(false_positive <- births_abx_sepsis_low_risk * (1 - sepsis_prev_low_risk) * (1 - specificity))
  
#' Has sepsis and not treated
(false_negative <- births_abx_sepsis_low_risk * sepsis_prev_low_risk * (1 - sensitivity))
  
#' Doesn't have sepsis and not treated
(true_negative <- births_abx_sepsis_low_risk * (1 - sepsis_prev_low_risk) * specificity)

stopifnot(true_positive + false_positive + false_negative + true_negative == births_abx_sepsis_low_risk)

#' Estimate that around half of false negatives would later be identified
harm_false_negative <- benefit_correct_sepsis_diagnosis * 0.5

benefit_test <- true_positive * benefit_correct_sepsis_diagnosis + true_negative * amr_burden_per_abx - harm_false_negative * false_negative - amr_burden_per_abx * false_positive
cost_test <- (cost_culture + total_cost_rdt) * births_abx_sepsis_low_risk + (true_positive + false_positive) * cost_treatment

#' Cost per DALY
(cost_per_daly_s2 <- cost_test / benefit_test)
```
